Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-27 manuscript.tex
section 3

% polychronization
Interestingly, the paper by~\citep{izhikevich_polychronization_2006} stirred a lively debate in the field of computational neuroscience, with a general positive acceptance, but relatively few works extended this seminal paper. Indeed, there were already existing models of synaptic delay learning in spiking neural networks, see for instance~\citep{huning_synaptic_1998} or~\citep{eurich_dynamics_1999} yet they had not shown potential applications to the detection of spiking motifs. A popular model for the detection of latency patterns is the \emph{tempotron}~\citep{gutig_tempotron_2006}. This model is in particular reviewed in~\citep{gutig_spike_2014}. The Tempotron is a supervised synaptic learning algorithm which classifies a distractor from a target motif. This aims at extending the perceptron which does not incorporate a spike timing framework. The tempotron learning rule is derived by an optimization process and takes the form of a STDP rule.  The limits of this model is that it's output is only binary and that it's storage capacity are limited. The recent ``multi-neuronal spike sequence detector'' architecture integrates the weight- and delay-adjustment methods by combining plasticity with the neurocomputational feature spike latency~\citep{susi_nmnsd-spiking_2021}. See also the extensive (graph-centric) review on synchronization in time-varying networks~\citep{ghosh_synchronized_2022}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Extensions of Izhikevitch's polychronization model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
An extension of~\citep{izhikevich_polychronization_2006} was made in a very detailed work aiming at reproducing the polychronization model~\citep{pauli_reproducing_2018}. Indeed, while the original paper contained material within the text to reproduce the whole model (using MATLAB), it was not complete for the reproduction of all results presented in that manuscript. This more recent work details how this code was slightly corrected. It comes with python code and a version control system detailing the whole process used to give provenance to the different steps in this scientific process.

Another recent work gives a Bayesian account in a similar model~\citep{guise_bayesian_2014}. Based on the fact that previous methods for studying polychronous neural groups (PNG) activation response to stimuli have been limited by the template-based methods used to identify PNG activation, the authors outline a new method that overcomes these difficulties by establishing a probabilistic interpretation of PNG activation. They demonstrate the use of this method by investigating the claim that PNGs might provide the foundation of a representational system. Stimulation of the trained network produces the activation of a PNG, ie the propagation of firing activity through multiple layers due to convergent patterns of firing.

@@

% Training delays ~\citep{state_training_2019}
A whole range of models of latency motifs detection are using STDP-based learning rules. For instance,~\citep{kheradpisheh_stdp-based_2018} implements a STDP-based spiking deep convolutional neural networks for object recognition;~\citep{tavanaei_representation_2018} develops a form of spike-based, competitive learning applied for unsupervised learning. In~\citep{russo_cell_2017}, the authors developed novel machine learning tools and statistical tests for unsupervised spatio-temporal pattern detection in non-stationary environments, which were applied to simultaneous electro-physiological recordings from tens to hundreds of neurons for decoding cognitive processes from neural activity.
%
Modified STPD rules have been used for synchronous coherence detection~\citep{perrinet_coherence_2002}, for the learning of specific receptive fields~\citep{perrinet_networks_2001} and were also extended to recurrent Neuronal Networks~\citep{gilson_stdp_2010} or delay selection~\citep{datadien_right_2011}, in particular for recurrent networks of spiking neurons receiving oscillatory inputs~\citep{kerr_delay_2013} which targets for the selective potentiation of recurrent connections with different axonal and dendritic delays during oscillatory activity. More generally, our ability to track and respond to rapidly changing visual stimuli, such as a fast-moving tennis ball, indicates that the brain is capable of extrapolating the trajectory of a moving object to predict its current position, despite the delays that result from neural transmission. Here, we show how the neural circuits underlying this ability can be learned through spike-timing-dependent synaptic plasticity and that these circuits emerge spontaneously and without supervision. This demonstrates how the neural transmission delays can, in part, be compensated to implement the extrapolation mechanisms required to predict where a moving object is at the present moment~\citep{burkitt_predictive_2021}.

Specifically, a recent work proposed a bio-plausible unsupervised delay learning for extracting temporal features in spiking neural networks~\citep{nadafian_bio-plausible_2020}. The authors provided some mathematical proofs to show that our learning rule gives a neuron the ability to learn repeating spatio-temporal patterns. Furthermore, the experimental results of applying an STDP-based spiking neural network was applied on Random Dot Kinematogram. Another model of synaptic delay-weight plasticity~\citep{zhang_supervised_2020} integrates synaptic delay plasticity into supervised learning and proposes a novel learning method that adjusts both the synaptic delays and weights of the learning neurons to make them fire precisely timed spikes. % Uses synthetic data to extend the existing Remote Supervised Method (ReSuMe) method.
In a recent paper~\citep{luo_supervised_2022}, authors propose a gradient descent-based learning algorithm for synaptic delays to enhance the sequential learning performance of a single spiking neuron. In this algorithm, information is encoded in the relative timing of individual neuronal spikes, and learning is performed based on the exact derivatives of the postsynaptic spike times with respect to presynaptic spike times.
In yet another computational model,~\citet{sun_learning_2016} show that the frequently activated polychronous neural groups can be learned by readout neurons with joint weight-delay spike-timing-dependent plasticity. This uses a joint Weight-Delay Spike-Timing-Dependent Plasticity and show an efficient learning of polychronous groups.

@@

%: >>>>>>>> Laurent is here <<<<<<<<<<<<<<
\subsection{Scaling up to very large scale data}
Over the past decade, tremendous technological advances across several disciplines have dramatically expanded the frontiers of experimentally accessible neuroscientific facts. Bridging across different spatial and temporal scales, combination of in vivo two photon imaging, large population recording-array technologies, optogenetic circuit control tools, transgenic manipulations as well as large volume circuit reconstructions are now used to examine the function, structure and dynamics of neural networks on an unprecedented level of detail and precision. The daunting complexity of the biological reality revealed by these technologies highlights the importance of neurophysics to provide a conceptual bridge between abstract principles of brain function and their biological implementations within neural circuits.

@@

% autres méthodes

>>>>

% polychronization
Interestingly, the paper by~\citep{izhikevich_polychronization_2006} stirred a lively debate in the field of computational neuroscience, with a general positive acceptance, but relatively few works extended this seminal paper. Indeed, there were already existing models of synaptic delay learning in spiking neural networks, see for instance~\citep{huning_synaptic_1998} or~\citep{eurich_dynamics_1999} yet they had not shown potential applications to the detection of spiking motifs. A popular model for the detection of latency patterns is the \emph{tempotron}~\citep{gutig_tempotron_2006}. This model is in particular reviewed in~\citep{gutig_spike_2014}. The Tempotron is a supervised synaptic learning algorithm which classifies a distractor from a target motif. This aims at extending the perceptron which does not incorporate a spike timing framework. The tempotron learning rule is derived by an optimization process and takes the form of a STDP rule.  The limits of this model is that it's output is only binary and that it's storage capacity are limited. The recent ``multi-neuronal spike sequence detector'' architecture integrates the weight- and delay-adjustment methods by combining plasticity with the neurocomputational feature spike latency~\citep{susi_nmnsd-spiking_2021}. See also the extensive (graph-centric) review on synchronization in time-varying networks~\citep{ghosh_synchronized_2022}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \subsection{Extensions of Izhikevitch's polychronization model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
An extension of~\citep{izhikevich_polychronization_2006} was made in a very detailed work aiming at reproducing the polychronization model~\citep{pauli_reproducing_2018}. Indeed, while the original paper contained material within the text to reproduce the whole model (using MATLAB), it was not complete for the reproduction of all results presented in that manuscript. This more recent work details how this code was slightly corrected. It comes with python code and a version control system detailing the whole process used to give provenance to the different steps in this scientific process.
%
Another recent work gives a Bayesian account in a similar model~\citep{guise_bayesian_2014}. Based on the fact that previous methods for studying polychronous neural groups (PNG) activation response to stimuli have been limited by the template-based methods used to identify PNG activation, the authors outline a new method that overcomes these difficulties by establishing a probabilistic interpretation of PNG activation. They demonstrate the use of this method by investigating the claim that PNGs might provide the foundation of a representational system. Stimulation of the trained network produces the activation of a PNG, ie the propagation of firing activity through multiple layers due to convergent patterns of firing.

@@

% Training delays ~\citep{state_training_2019}
A whole range of models of latency motifs detection are using STDP-based learning rules. For instance,~\citep{kheradpisheh_stdp-based_2018} implements a STDP-based spiking deep convolutional neural networks for object recognition;~\citep{tavanaei_representation_2018} develops a form of spike-based, competitive learning applied for unsupervised learning. 
%
Modified STPD rules have been used for synchronous coherence detection~\citep{perrinet_coherence_2002}, for the learning of specific receptive fields~\citep{perrinet_networks_2001} and were also extended to recurrent Neuronal Networks~\citep{gilson_stdp_2010} or delay selection~\citep{datadien_right_2011}. In particular, this has been applied for recurrent networks of spiking neurons receiving oscillatory inputs~\citep{kerr_delay_2013} which targets for the selective potentiation of recurrent connections with different axonal and dendritic delays during oscillatory activity. More generally, our ability to track and respond to rapidly changing visual stimuli, such as a fast-moving tennis ball, indicates that the brain is capable of extrapolating the trajectory of a moving object to predict its current position, despite the delays that result from neural transmission. Here, we show how the neural circuits underlying this ability can be learned through spike-timing-dependent synaptic plasticity and that these circuits emerge spontaneously and without supervision. This demonstrates how the neural transmission delays can, in part, be compensated to implement the extrapolation mechanisms required to predict where a moving object is at the present moment~\citep{burkitt_predictive_2021}.

Specifically, a recent work proposed a bio-plausible unsupervised delay learning for extracting temporal features in spiking neural networks~\citep{nadafian_bio-plausible_2020}. The authors provided some mathematical proofs to show that our learning rule gives a neuron the ability to learn repeating spatio-temporal patterns. Furthermore, the experimental results of applying an STDP-based spiking neural network was applied on Random Dot Kinematogram. Another model of synaptic delay-weight plasticity~\citep{zhang_supervised_2020} integrates synaptic delay plasticity into supervised learning and proposes a novel learning method that adjusts both the synaptic delays and weights of the learning neurons to make them fire precisely timed spikes. 
This was also presented by~\citep{wang_delay_2019} which propose a supervised delay learning algorithm for spiking neurons with temporal encoding, in which both the weight and delay of a synaptic connection can be adjusted to enhance the learning performance.
% Uses synthetic data to extend the existing Remote Supervised Method (ReSuMe) method.
In a recent paper~\citep{luo_supervised_2022}, authors propose a gradient descent-based learning algorithm for synaptic delays to enhance the sequential learning performance of a single spiking neuron. In this algorithm, information is encoded in the relative timing of individual neuronal spikes, and learning is performed based on the exact derivatives of the postsynaptic spike times with respect to presynaptic spike times.
In yet another computational model,~\citet{sun_learning_2016} show that the frequently activated polychronous neural groups can be learned by readout neurons with joint weight-delay spike-timing-dependent plasticity. This uses a joint Weight-Delay Spike-Timing-Dependent Plasticity and show an efficient learning of polychronous groups.

@@

%: >>>>>>>> Laurent is here <<<<<<<<<<<<<<
\subsection{Scaling up to very large scale data}

In~\citep{russo_cell_2017}, the authors developed novel machine learning tools and statistical tests for unsupervised spatio-temporal pattern detection in non-stationary environments, which were applied to simultaneous electro-physiological recordings from tens to hundreds of neurons for decoding cognitive processes from neural activity.

Over the past decade, tremendous technological advances across several disciplines have dramatically expanded the frontiers of experimentally accessible neuroscientific facts. Bridging across different spatial and temporal scales, combination of in vivo two photon imaging, large population recording-array technologies, optogenetic circuit control tools, transgenic manipulations as well as large volume circuit reconstructions are now used to examine the function, structure and dynamics of neural networks on an unprecedented level of detail and precision. The daunting complexity of the biological reality revealed by these technologies highlights the importance of neurophysics to provide a conceptual bridge between abstract principles of brain function and their biological implementations within neural circuits.

@@

% autres méthodes
spatitemporal kernels: \citep{yu_stsc-snn_2022}
